Navigating Hostility: The Effect of Nonprofit Transparency and Accountability on Donor Preferences in the Face of Shrinking Civic Space

### 2025-07-02 manuscript/manuscript.qmd
Final edits

While INGOs have control over organization-level factors that determine their effectiveness, the political environment in the host countries INGOs work in---including the restrictions on civil society that cause many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do individual donors in the US feel about donating to legally besieged NGOs abroad? How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? While there are many organization-level factors that can determine an NGO's impact, including organizational structure, financial transparency, accountability, leadership, and strategy [@MitchellSchmitzBruno-vanVijfeijken:2020], in this paper, we focus on the two factors that are most visible to donors outside of the organizations and on which information is available publicly---financial transparency and accountability.

@@

We fielded a conjoint experiment to a nationally representative sample of `r details$n_nice` adults in the US. This experiment allows us to assess individual-level giving preferences among likely donors in the US towards NGOs abroad working in a variety of issue areas---human rights, environment, emergency response, and refugee relief.[^not-elite] More importantly, it allows us to directly compare the relative impact of host government restriction of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism of and restrictions on NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect.

@@

But as INGOs gained more power in global policy circles, they faced a number of criticisms. Many organizations struggle to balance efficient large-scale operations with the grassroots connections and consensus-building that drove their early success [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008]. As a result, many INGOs worked to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organization-level factors that can potentially impact nonprofit efficacy, including organizational and governance structure, leadership, strategy, transparency, and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries and devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy involves whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning, and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organization-level factors) are often not visible to the average individual donor. We therefore focus on two organization-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, donors rely on cues and heuristics when deciding to support an NGO. Compared to other organization-level factors, news about financial mismanagement and lack of transparency and accountability may be more easily available. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the United Nations' Oil-for-Food program, aimed at alleviated Iraq's sanctions-induced hardships in the 1990s, which was marred by revelations of widespred financial fraud, limited food distribution, and lack of accountability in 2004. Numerous other cases [see @GibelmanGelman:2001 and @GibelmanGelman:2004] demonstrate how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust and reduce donations to nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watchdogs such as Charity Navigator often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

Thus, in this paper, we explore the effects of two possible categories of heuristic frames that can impact individual-level giving: (1) organization-level heuristics, reflected in practices of transparency and accountability, and (2) structural heuristics, particularly an NGO's relationship with its host country. We hypothesize that each type plays a role in shaping individual preferences for engaging in philanthropy.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways, such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@VaccaroMadsen:2006].

@@

Accountability also influences donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also reflects how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders seek assurance that their donations will make a real difference---effective accountability can encourage NGOs to become more closely aligned with community practices and help expand stakeholder support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, deteriorated reputation, and lower perceived quality [@Becker:2018].

Though individual donors would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead [@HungHagerTian:2023], individuals interested in supporting charities cannot thoroughly vet every aspect of the organization [@Tremblay-BoirePrakash:2017; @ChaudhryHeiss:2021]. To more easily judge financial transparency and accountability, donors may use ratings by watchdog organizations like Charity Navigator and GuideStar as heuristics to guide their decisions.[^cn-multiple] Research shows that initiatives to address information gaps by these organizations can substantially increase donations [@Sloan:2009; @GordonKnockNeely:2009; @Tremblay-BoirePrakash:2017]. We expect that donors will respond to signals that organizations engage in transparency and accountability practices:

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as the Russian Foreign Agent law, are not grounded in international legal principles and do not give nonprofits the freedom of association or access to resources [@Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

Some NGOs have adapted to these legal crackdowns by recreating their organizations and changing their mission to avoid directly confronting the government. Research on anti-NGO laws in Bangladesh and Zambia shows that NGOs with broad missions shifted from advocacy to service work, while those focused mainly on transnational advocacy altered their targets, issues, and language [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer organizations work on contentious causes, instead focusing on safer issues like health and education. These changes in the civil society legal environment have influenced official donor responses. While multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022].

However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the ground [@EasterlyWilliamson:2011]. When a potential recipient NGO faces legal crackdown in its host country, private individual donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for perhaps doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryHeiss:2021 find conflicting donor responses to foreign restrictions---while some prospective donors punish nonprofits for doing something "wrong" to incur host government restrictions, most tended to increase their support, stating that restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust---i.e., those who frequently volunteer, regularly donate to charity, and trust political institutions---are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. Thus, anti-NGO laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and seek to restrict their work.

This is especially true for nonprofits that rely more heavily on private funds, rather than government funds. Previous research finds that, "individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds" [@ChaudhryHeiss:2021, 496]. Similar trends can be seen even after nonprofits face restrictive political and civic environments in other specific issue areas. For instance, in 2022, after the Supreme Court overturned Roe v. Wade, private philanthropy towards abortion funds skyrocketed [@PBSNews:2022]---a significant development as abortion funds rely largely on individual donors [@Cohen:2022]. Therefore, we expect that anti-NGO restrictions imposed by host-country governments are likely to boost individual donors' inclination to donate to affected organizations.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Moreover, states pass anti-civil society laws in response to broader political trends within their borders, and they regulate NGOs strategically to maximize the benefits and minimize the costs of working with international NGOs [@Heiss:2017]. States are more likely to restrict NGOs when organizations pose a threat to regime stability or preferences---for instance, when INGO issue areas threaten government policies, or when INGOs receive substantial foreign funding [@DupuyRonPrakash:2015].

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018]. Following screening, we received `r details$n_nice` viable responses. Tables A1, A2, and A3 provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity. Additionally, Table A4 compares sample characteristics with nationally representative estimates from the 2019 Current Population Survey (CPS). Our sample matches national proportions of age, marital status, and education, and contains slightly more men and slightly higher income levels than the general population. Due to the initial screening question that limited participants to those with an interest in charitable activity, respondents are far more likely to have donated, volunteered, or voted in the past year than population CPS estimates. Our reported causal effects are therefore not representative of the entire population and instead are generalizable to people interested in charitable giving.

@@

After collecting baseline information on respondent demographics and attitudes toward charity and voluntarism, we showed participants a set of randomly shuffled hypothetical international NGOs described with randomly shuffled features or attributes. In this paper, we hypothesize and empirically test the effect of financial transparency, accountability, and host country relationship on the likelihood of donations. In our experiment, we included a few additional treatments to aid with the interpretation and comparison of effect sizes for our treatments of interest. In total, we varied six different organizational and structural attributes that might have an effect on donor behavior: (1) organization name and branding, (2) organization issue area, (3) financial transparency practices, (4) accountability practices, (5) funding sources, and (6) relationship with host government (see @tbl-design).

@@

We analyze the results using a multilevel Bayesian multinomial model (see the appendix for complete model details). Our experimental data has a natural hierarchical structure, with 3 questions nested inside 12 separate experimental tasks, nested inside each of the `r details$n_nice` respondents, which lends itself to multilevel modeling [@JensenMarbleScheve:2021]. Since it was impossible for every respondent to see every possible combination of all 12,000+ experimental tasks, multilevel modeling allows us to pool together information about respondents with similar characteristics facing similar sets of choices. Moreover, using random respondent effects provides natural regularization and shrinkage for our estimates---experimental tasks that happened to appear more often due to chance will be accounted for and their frequency will not bias the overall causal effect. We define our model and priors in @eq-model.

@@

Having explored the effects of general organizational characteristics on the propensity to donate, we can test our first hypotheses and examine the effects of our treatments of interest. As seen in the marginal means in @fig-all-results, both transparency and accountability are strong signals of organizational deservingness. Respondents strongly prefer organizations that engage in either transparency or accountability---both treatments have a posterior marginal mean of roughly 30% compared to a baseline equally-at-random probability of 25% (transparency: `r mms$transparency_yes$nice_simple`; accountability: `r mms$accountability_yes$nice_simple`). The AMCEs for each treatment show a roughly 10 percentage point increase in the probability of selection compared to organizations that do not engage in transparency or accountability (transparency: `r amce$transparency_yes$nice_full`; accountability: `r amce$accountability_yes$nice_full`). This effect is nearly the same order of magnitude as the Red Cross effect---an NGO that chooses to be more transparent or that takes steps to demonstrate greater accountability can expect to see an increase in the probability of selection equivalent to the boost of the brand name effect associated with the Red Cross. As predicted, we thus find strong support for both H~1a~ and H~1b~: if NGOs are financially transparent or engage in accountability practices, then individual private donors are roughly 10 percentage points more likely to donate to them. Donors appear to reward NGOs for their efforts to disclose their funding and be more transparent.

There is also some evidence that these two effects are equivalent and perhaps interchangeable. @fig-transp-acc (and Table A8) shows the posterior marginal means for all four combinations of possible levels of transparency and accountability. When respondents see both as "no" simultaneously, the estimated marginal mean is lower than any of the individual treatments in @fig-all-results, at `r mms_ta$transparency_no_accountability_no$pct` (`r mms_ta$transparency_no_accountability_no$nice_ci`), or 9 percentage points lower than the baseline probability of 25%. When either transparency or accountability is set to "yes", the estimated marginal mean is essentially identical at `r mms$transparency_yes$pct` and `r mms$accountability_yes$pct`, respectively. When both treatments are set to "yes", the estimated marginal mean is `r mms_ta$transparency_yes_accountability_yes$pct` (`r mms_ta$transparency_yes_accountability_yes$nice_ci`), which is roughly the same as the overall estimated marginal mean for the Red Cross. Holding all other treatments constant, the effect of transparency and accountability practices on their own are generally the same, and when combined, the overall favorability of the organization increases substantially. This could indicate that donors do not care about which specific type of organizational practice an NGO engages in and that instead they are looking for some sort of signal that the organization is following best practices in transparency or accountability.

@@

To test our second hypothesis regarding structural heuristics, we explore the effect of an NGO's host country conditions on the propensity to donate. Contrary to our expectations, respondents appear to prefer donating to NGOs with friendly host government relationships, with a marginal mean of `r mms$friendly_relationship_with_government$pct` (compared to a baseline of 25%; `r mms$friendly_relationship_with_government$nice_ci`). Respondents are less likely to donate when an organization is criticized by its host government, and far less likely when an organization faces restrictions. Compared to other treatments, an organization facing restrictions elicits a similar negative preference among potential donors as an organization without transparency or accountability measures. However, an organization with a friendly relationship with the host government leads potential donors to perceive it as equivalent to engaging in transparency or accountability.

@@

Our results also have multiple important theoretical implications that should be explored in future research. First, future research should test how well these results map on to elite individual donors and foundations---not just the average private donor. As large foundations are more likely to publish data about their giving than the average private donor, it may be possible to determine how the changes in preferences elicited by these heuristics translate into actual donation behavior. Second, it is also important to test how these findings translate to non-American populations. Due to structural differences in the European nonprofit sector, where NGOs rely less on government funding and more on private funding [@Stroup:2012], donors are likely motivated by different concerns. Third, many states use negative rhetoric designed to sow mistrust between these groups and communities that support them. While donor responses have largely focused on navigating and adapting to legal restrictions, the use of such rhetoric raises broader questions about research and policy in the nonprofit sector: how is such rhetoric changing public attitudes and donations towards these groups? Are the effects of transparency and accountability found in this paper still likely to hold true if the home country also starts smearing these NGOs? Finally, though this paper examined legal crackdowns on INGOs, this issue is even more challenging for local NGOs in Global South countries, where individual donors might avoid giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to unfavorable tax benefits or fear of retribution [@Brechenmacher:2017; @DupuyRonPrakash:2015]. Future research should examine how local NGOs can overcome individual foreign donors' concerns about host governments' criticisms of and crackdown on these groups.

>>>>

While INGOs have control over organization-level factors that determine their effectiveness, the political environment in the host countries INGOs work in---including the restrictions on civil society that cause many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do individual donors in the US feel about donating to legally besieged NGOs abroad? How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? While there are many organization-level factors that can determine an NGO's impact, including organizational structure, financial transparency, accountability, leadership, and strategy [@MitchellSchmitzBruno-vanVijfeijken:2020], in this article, we focus on the two factors that are most visible to donors outside of the organizations and on which information is available publicly---financial transparency and accountability.

@@

We fielded a conjoint experiment to a nationally representative sample of `r details$n_nice` adults in the US. This experiment allows us to assess individual-level giving preferences among likely donors in the US toward NGOs abroad working in a variety of issue areas---human rights, environment, emergency response, and refugee relief.[^not-elite] More importantly, it allows us to directly compare the relative impact of host government restriction of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism of and restrictions on NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect.

@@

But as INGOs gained more power in global policy circles, they faced a number of criticisms. Many organizations struggle to balance efficient large-scale operations with the grassroots connections and consensus-building that drove their early success [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008]. As a result, many INGOs worked to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organization-level factors that can potentially affect nonprofit efficacy, including organizational and governance structure, leadership, strategy, transparency, and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries and devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy involves whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning, and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organization-level factors) are often not visible to the average individual donor. We therefore focus on two organization-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, donors rely on cues and heuristics when deciding to support an NGO. Compared with other organization-level factors, news about financial mismanagement and lack of transparency and accountability may be more easily available. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the United Nations' Oil-for-Food program, aimed at alleviated Iraq's sanctions-induced hardships in the 1990s, which was marred by revelations of widespread financial fraud, limited food distribution, and lack of accountability in 2004. Numerous other cases [see @GibelmanGelman:2001; @GibelmanGelman:2004] demonstrate how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust and reduce donations to nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watchdogs such as Charity Navigator often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

Thus, in this article, we explore the effects of two possible categories of heuristic frames that can affect individual-level giving: (1) organization-level heuristics, reflected in practices of transparency and accountability, and (2) structural heuristics, particularly an NGO's relationship with its host country. We hypothesize that each type plays a role in shaping individual preferences for engaging in philanthropy.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities" [@VaccaroMadsen:2006, p. 146], is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways, such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@VaccaroMadsen:2006].

@@

Accountability also influences donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also reflects how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders seek assurance that their donations will make a real difference---effective accountability can encourage NGOs to become more closely aligned with community practices and help expand stakeholder support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgments or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, deteriorated reputation, and lower perceived quality [@Becker:2018].

Although individual donors would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead [@HungHagerTian:2023], individuals interested in supporting charities cannot thoroughly vet every aspect of the organization [@Tremblay-BoirePrakash:2017; @ChaudhryHeiss:2021]. To more easily judge financial transparency and accountability, donors may use ratings by watchdog organizations like Charity Navigator and GuideStar as heuristics to guide their decisions.[^cn-multiple] Research shows that initiatives to address information gaps by these organizations can substantially increase donations [@Sloan:2009; @GordonKnockNeely:2009; @Tremblay-BoirePrakash:2017]. We expect that donors will respond to signals that organizations engage in transparency and accountability practices:

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this article, such as the Russian Foreign Agent law, are not grounded in international legal principles and do not give nonprofits the freedom of association or access to resources [@Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

Some NGOs have adapted to these legal crackdowns by recreating their organizations and changing their mission to avoid directly confronting the government. Research on anti-NGO laws in Bangladesh and Zambia shows that NGOs with broad missions shifted from advocacy to service work, while those focused mainly on transnational advocacy altered their targets, issues, and language [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer organizations work on contentious causes, instead focusing on safer issues like health and education. These changes in the civil society legal environment have influenced official donor responses. While multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Furthermore, advocacy-oriented donors---that is,, those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022].

However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore may be in a better position to respond to recipients' needs on the ground [@EasterlyWilliamson:2011]. When a potential recipient NGO faces legal crackdown in its host country, private individual donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for perhaps doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryHeiss:2021 find conflicting donor responses to foreign restrictions---while some prospective donors punish nonprofits for doing something "wrong" to incur host government restrictions, most tended to increase their support, stating that restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust---that is, those who frequently volunteer, regularly donate to charity, and trust political institutions---are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. Thus, anti-NGO laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and seek to restrict their work.

This is especially true for nonprofits that rely more heavily on private funds, rather than government funds. Previous research finds that, "individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds" [@ChaudhryHeiss:2021, 496]. Similar trends can be seen even after nonprofits face restrictive political and civic environments in other specific issue areas. For instance, in 2022, after the Supreme Court overturned Roe v. Wade, private philanthropy toward abortion funds skyrocketed [@PBSNews:2022]---a significant development as abortion funds rely largely on individual donors [@Cohen:2022]. Therefore, we expect that anti-NGO restrictions imposed by host-country governments are likely to boost individual donors' inclination to donate to affected organizations.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior because (1) NGOs have less direct control over their political environments and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Moreover, states pass anti-civil society laws in response to broader political trends within their borders, and they regulate NGOs strategically to maximize the benefits and minimize the costs of working with international NGOs [@Heiss:2017]. States are more likely to restrict NGOs when organizations pose a threat to regime stability or preferences---for instance, when INGO issue areas threaten government policies, or when INGOs receive substantial foreign funding [@DupuyRonPrakash:2015].

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018]. Following screening, we received `r details$n_nice` viable responses. Tables A1, A2, and A3 provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity. In addition, Table A4 compares sample characteristics with nationally representative estimates from the 2019 Current Population Survey (CPS). Our sample matches national proportions of age, marital status, and education, and contains slightly more men and slightly higher income levels than the general population. Due to the initial screening question that limited participants to those with an interest in charitable activity, respondents are far more likely to have donated, volunteered, or voted in the past year than population CPS estimates. Our reported causal effects are therefore not representative of the entire population and instead are generalizable to people interested in charitable giving.

@@

After collecting baseline information on respondent demographics and attitudes toward charity and voluntarism, we showed participants a set of randomly shuffled hypothetical international NGOs described with randomly shuffled features or attributes. In this article, we hypothesize and empirically test the effect of financial transparency, accountability, and host country relationship on the likelihood of donations. In our experiment, we included a few additional treatments to aid with the interpretation and comparison of effect sizes for our treatments of interest. In total, we varied six different organizational and structural attributes that might have an effect on donor behavior: (1) organization name and branding, (2) organization issue area, (3) financial transparency practices, (4) accountability practices, (5) funding sources, and (6) relationship with host government (see @tbl-design).

@@

We analyze the results using a multilevel Bayesian multinomial model (see the appendix for complete model details). Our experimental data has a natural hierarchical structure, with 3 questions nested inside 12 separate experimental tasks, nested inside each of the `r details$n_nice` respondents, which lends itself to multilevel modeling [@JensenMarbleScheve:2021]. As it was impossible for every respondent to see every possible combination of all 12,000+ experimental tasks, multilevel modeling allows us to pool together information about respondents with similar characteristics facing similar sets of choices. Moreover, using random respondent effects provides natural regularization and shrinkage for our estimates---experimental tasks that happened to appear more often due to chance will be accounted for and their frequency will not bias the overall causal effect. We define our model and priors in @eq-model.

@@

Having explored the effects of general organizational characteristics on the propensity to donate, we can test our first hypotheses and examine the effects of our treatments of interest. As seen in the marginal means in @fig-all-results, both transparency and accountability are strong signals of organizational deservingness. Respondents strongly prefer organizations that engage in either transparency or accountability---both treatments have a posterior marginal mean of roughly 30% compared with a baseline equally-at-random probability of 25% (transparency: `r mms$transparency_yes$nice_simple`; accountability: `r mms$accountability_yes$nice_simple`). The AMCEs for each treatment show a roughly 10 percentage point increase in the probability of selection compared to organizations that do not engage in transparency or accountability (transparency: `r amce$transparency_yes$nice_full`; accountability: `r amce$accountability_yes$nice_full`). This effect is nearly the same order of magnitude as the Red Cross effect---an NGO that chooses to be more transparent or that takes steps to demonstrate greater accountability can expect to see an increase in the probability of selection equivalent to the boost of the brand name effect associated with the Red Cross. As predicted, we thus find strong support for both H~1a~ and H~1b~: if NGOs are financially transparent or engage in accountability practices, then individual private donors are roughly 10 percentage points more likely to donate to them. Donors appear to reward NGOs for their efforts to disclose their funding and be more transparent.

There is also some evidence that these two effects are equivalent and perhaps interchangeable. @fig-transp-acc (and Table A8) shows the posterior marginal means for all four combinations of possible levels of transparency and accountability. When respondents see both as "no" simultaneously, the estimated marginal mean is lower than any of the individual treatments in @fig-all-results, at `r mms_ta$transparency_no_accountability_no$pct` (`r mms_ta$transparency_no_accountability_no$nice_ci`), or 9 percentage points lower than the baseline probability of 25%. When either transparency or accountability is set to "yes," the estimated marginal mean is essentially identical at `r mms$transparency_yes$pct` and `r mms$accountability_yes$pct`, respectively. When both treatments are set to "yes," the estimated marginal mean is `r mms_ta$transparency_yes_accountability_yes$pct` (`r mms_ta$transparency_yes_accountability_yes$nice_ci`), which is roughly the same as the overall estimated marginal mean for the Red Cross. Holding all other treatments constant, the effect of transparency and accountability practices on their own are generally the same, and when combined, the overall favorability of the organization increases substantially. This could indicate that donors do not care about which specific type of organizational practice an NGO engages in and that instead they are looking for some sort of signal that the organization is following best practices in transparency or accountability.

@@

To test our second hypothesis regarding structural heuristics, we explore the effect of an NGO's host country conditions on the propensity to donate. Contrary to our expectations, respondents appear to prefer donating to NGOs with friendly host government relationships, with a marginal mean of `r mms$friendly_relationship_with_government$pct` (compared with a baseline of 25%; `r mms$friendly_relationship_with_government$nice_ci`). Respondents are less likely to donate when an organization is criticized by its host government, and far less likely when an organization faces restrictions. Compared with other treatments, an organization facing restrictions elicits a similar negative preference among potential donors as an organization without transparency or accountability measures. However, an organization with a friendly relationship with the host government leads potential donors to perceive it as equivalent to engaging in transparency or accountability.

@@

Our results also have multiple important theoretical implications that should be explored in future research. First, future research should test how well these results map on to elite individual donors and foundations---not just the average private donor. As large foundations are more likely to publish data about their giving than the average private donor, it may be possible to determine how the changes in preferences elicited by these heuristics translate into actual donation behavior. Second, it is also important to test how these findings translate to non-American populations. Due to structural differences in the European nonprofit sector, where NGOs rely less on government funding and more on private funding [@Stroup:2012], donors are likely motivated by different concerns. Third, many states use negative rhetoric designed to sow mistrust between these groups and communities that support them. While donor responses have largely focused on navigating and adapting to legal restrictions, the use of such rhetoric raises broader questions about research and policy in the nonprofit sector: how is such rhetoric changing public attitudes and donations towards these groups? Are the effects of transparency and accountability found in this article still likely to hold true if the home country also starts smearing these NGOs? Finally, though this article examined legal crackdowns on INGOs, this issue is even more challenging for local NGOs in Global South countries, where individual donors might avoid giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to unfavorable tax benefits or fear of retribution [@Brechenmacher:2017; @DupuyRonPrakash:2015]. Future research should examine how local NGOs can overcome individual foreign donors' concerns about host governments' criticisms of and crackdown on these groups.
